\subsection{Agent State Machine Formalization}
\label{sec:state-machine}

This section provides a formal specification of the contagion state machine that governs agent health transitions. The model extends classical SIR compartmental dynamics \cite{kermack1927contribution} with an additional DECEASED state to capture fatality outcomes, following the SIRD framework used in recent COVID-19 modeling literature \cite{fanelli2020new, calafiore2020modified}.

\subsubsection{State Definitions}

The contagion simulation defines a finite state machine with four mutually exclusive health states:

\begin{enumerate}
    \item \textbf{HEALTHY} ($S$): Susceptible agents with no active infection. Agents in this state can contract the contagion through proximity contact or floor contamination exposure.
    \item \textbf{INFECTED} ($I$): Agents with active infection capable of transmitting the contagion. This state has a deterministic duration governed by the recovery time parameter.
    \item \textbf{RECOVERED} ($R$): Post-infection agents with temporary immunity. Cannot be reinfected until immunity expires.
    \item \textbf{DECEASED} ($D$): Terminal state representing fatality. Movement is disabled and no further state transitions occur.
\end{enumerate}

The state space is formally defined as:
\begin{equation}
\label{eq:state-space}
\mathcal{S} = \{S, I, R, D\}
\end{equation}

where each agent $a_i$ maintains a current state $s_i \in \mathcal{S}$ at time $t$.

\subsubsection{State Transition Diagram}

Figure~\ref{fig:state-machine} illustrates the complete state transition diagram with all valid transitions and their triggering conditions.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=2.5cm and 3cm,
    state/.style={circle, draw, thick, minimum size=1.5cm, align=center, font=\small\bfseries},
    arrow/.style={->, >=stealth, thick, font=\footnotesize},
    every edge/.style={draw, arrow}
]
    % States
    \node[state, fill=green!20] (healthy) {HEALTHY\\($S$)};
    \node[state, fill=red!20, right=of healthy] (infected) {INFECTED\\($I$)};
    \node[state, fill=blue!20, below=of healthy] (recovered) {RECOVERED\\($R$)};
    \node[state, fill=gray!30, below=of infected] (deceased) {DECEASED\\($D$)};
    
    % Transitions
    \draw[arrow] (healthy) -- node[above] {$P_{\text{inf}}$} (infected);
    \draw[arrow] (infected) -- node[right] {$1 - P_{\text{fat}}$} (recovered);
    \draw[arrow] (infected) -- node[right, pos=0.3] {$P_{\text{fat}}$} (deceased);
    \draw[arrow] (recovered) -- node[left] {$P_{\text{expire}}$} (healthy);
    
    % Self-loop for INFECTED duration
    \draw[arrow] (infected) to[loop above] node {$\Delta t_{\text{rec}}$} (infected);
    
\end{tikzpicture}
\caption{Agent health state machine with transition probabilities. HEALTHY agents become INFECTED through proximity or floor contact ($P_{\text{inf}}$). INFECTED agents transition to RECOVERED with probability $(1 - P_{\text{fat}})$ or DECEASED with probability $P_{\text{fat}}$ after recovery time $\Delta t_{\text{rec}}$. RECOVERED agents return to HEALTHY when immunity expires.}
\label{fig:state-machine}
\end{figure}

\subsubsection{Transition Conditions and Probabilities}

Each state transition is governed by probabilistic conditions that ensure frame-rate independence. We denote the base per-second probability as $P_{\text{base}}$ and the frame delta time as $\Delta t$ (milliseconds).

\paragraph{HEALTHY $\to$ INFECTED Transition}

A HEALTHY agent transitions to INFECTED when exposed to either: (1) proximity contact with an infected agent, or (2) contaminated floor tile. The scaled probability for frame-rate independence is:

\begin{equation}
\label{eq:infection-prob}
P_{\text{inf}} = 1 - (1 - P_{\text{base}})^{\Delta t / 1000}
\end{equation}

For proximity transmission, the condition requires:
\begin{equation}
\label{eq:proximity-cond}
\|p_i - p_j\|_2 \leq r_{\text{inf}} \land s_j = I
\end{equation}

where $p_i, p_j$ are agent positions, $r_{\text{inf}}$ is the infection radius (default: 80 pixels), and $s_j = I$ indicates agent $j$ is infected.

For floor contamination transmission:
\begin{equation}
\label{eq:floor-cond}
a_i \in \text{AABB}(f_k) \land \ell_k > 0
\end{equation}

where $\text{AABB}(f_k)$ is the axis-aligned bounding box of floor tile $f_k$, and $\ell_k \in \{0, 1, 2, 3\}$ is the contamination level. The floor infection probability scales with level:
\begin{equation}
\label{eq:floor-prob}
P_{\text{floor}} = P_{\text{base}}^{\text{floor}} \times (\ell_k / 3)
\end{equation}

\paragraph{INFECTED $\to$ RECOVERED/DECEASED Transition}

After recovery time $\Delta t_{\text{rec}}$ (default: 15000 ms), an infected agent rolls for fatality:

\begin{equation}
\label{eq:fatality-cond}
s_i = I \land (t_{\text{now}} - t_{\text{inf},i}) \geq \Delta t_{\text{rec}} \Rightarrow 
\begin{cases}
s_i \leftarrow R & \text{with prob } (1 - P_{\text{fat}})\\
s_i \leftarrow D & \text{with prob } P_{\text{fat}}
\end{cases}
\end{equation}

where $t_{\text{inf},i}$ is the infection timestamp for agent $i$, and $P_{\text{fat}}$ is the fatality probability (default: 0.05).

\paragraph{RECOVERED $\to$ HEALTHY Transition}

Recovered agents lose immunity after duration $\Delta t_{\text{imm}}$ (default: 30000 ms):

\begin{equation}
\label{eq:immunity-expire}
s_i = R \land (t_{\text{now}} - t_{\text{rec},i}) \geq \Delta t_{\text{imm}} \Rightarrow s_i \leftarrow S
\end{equation}

where $t_{\text{rec},i}$ is the recovery timestamp. The transition probability $P_{\text{expire}} = 1$ when the time condition is satisfied.

\subsubsection{Floor Contamination Decay Model}

Floor tiles maintain contamination levels $\ell \in \{0, 1, 2, 3\}$ that decay over time. The decay follows a deterministic step function:

\begin{equation}
\label{eq:decay}
\ell_k(t + \Delta t_{\text{decay}}) = \max(0, \ell_k(t) - 1)
\end{equation}

where $\Delta t_{\text{decay}}$ is the decay interval (default: 10000 ms). Each infected agent occupying a floor tile increments its contamination level:

\begin{equation}
\label{eq:contaminate}
\ell_k \leftarrow \min(3, \ell_k + 1) \quad \text{if } a_i \in \text{AABB}(f_k) \land s_i = I
\end{equation}

Contamination levels map to visual overlays via color gradients:
\begin{itemize}
    \item \textbf{Level 0}: Clean (green overlay, $P_{\text{floor}} = 0$)
    \item \textbf{Level 1}: Low contamination (yellow overlay, $P_{\text{floor}} = P_{\text{base}}^{\text{floor}} / 3$)
    \item \textbf{Level 2}: Medium contamination (orange overlay, $P_{\text{floor}} = 2P_{\text{base}}^{\text{floor}} / 3$)
    \item \textbf{Level 3}: High contamination (red overlay, $P_{\text{floor}} = P_{\text{base}}^{\text{floor}}$)
\end{itemize}

\subsubsection{updateContagion Algorithm}

Algorithm~\ref{alg:update-contagion} presents the pseudocode for the main simulation loop that updates agent states and floor contamination each frame.

\begin{algorithm}[t]
\caption{updateContagion($\Delta t$)}
\label{alg:update-contagion}
\begin{algorithmic}[1]
\REQUIRE Frame delta time $\Delta t$ (ms)
\STATE $t_{\text{now}} \leftarrow t_{\text{now}} + \Delta t$
\STATE $P_{\text{scaled}} \leftarrow 1 - (1 - P_{\text{base}})^{\Delta t / 1000}$
\FOR{each agent $a_i$ with $s_i = S$}
    \STATE $exposed \leftarrow \text{false}$
    \FOR{each agent $a_j$ with $s_j = I$}
        \IF{$\|p_i - p_j\|_2 \leq r_{\text{inf}}$}
            \IF{$\text{random}() < P_{\text{scaled}}$}
                \STATE $exposed \leftarrow \text{true}$
                \STATE \textbf{break}
            \ENDIF
        \ENDIF
    \ENDFOR
    \IF{$\neg exposed$}
        \FOR{each floor tile $f_k$ with $\ell_k > 0$}
            \IF{$a_i \in \text{AABB}(f_k)$}
                \STATE $P_{\text{floor}} \leftarrow P_{\text{base}}^{\text{floor}} \times (\ell_k / 3)$
                \IF{$\text{random}() < 1 - (1 - P_{\text{floor}})^{\Delta t / 1000}$}
                    \STATE $exposed \leftarrow \text{true}$
                    \STATE \textbf{break}
                \ENDIF
            \ENDIF
        \ENDFOR
    \ENDIF
    \IF{$exposed$}
        \STATE $s_i \leftarrow I$, $t_{\text{inf},i} \leftarrow t_{\text{now}}$
    \ENDIF
\ENDFOR
\FOR{each agent $a_i$ with $s_i = I$}
    \IF{$(t_{\text{now}} - t_{\text{inf},i}) \geq \Delta t_{\text{rec}}$}
        \IF{$\text{random}() < P_{\text{fat}}$}
            \STATE $s_i \leftarrow D$, disable movement
        \ELSE
            \STATE $s_i \leftarrow R$, $t_{\text{rec},i} \leftarrow t_{\text{now}}$
        \ENDIF
    \ENDIF
\ENDFOR
\FOR{each agent $a_i$ with $s_i = R$}
    \IF{$(t_{\text{now}} - t_{\text{rec},i}) \geq \Delta t_{\text{imm}}$}
        \STATE $s_i \leftarrow S$
    \ENDIF
\ENDFOR
\FOR{each floor tile $f_k$}
    \IF{$(t_{\text{now}} - t_{\text{decay},k}) \geq \Delta t_{\text{decay}}$}
        \STATE $\ell_k \leftarrow \max(0, \ell_k - 1)$
        \STATE $t_{\text{decay},k} \leftarrow t_{\text{now}}$
    \ENDIF
    \FOR{each agent $a_i$ with $s_i = I$}
        \IF{$a_i \in \text{AABB}(f_k)$}
            \STATE $\ell_k \leftarrow \min(3, \ell_k + 1)$
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

The algorithm operates in five phases: (1) proximity infection checks for HEALTHY agents, (2) floor contamination checks for unexposed HEALTHY agents, (3) recovery/fatality rolls for INFECTED agents, (4) immunity expiry for RECOVERED agents, and (5) floor contamination decay and increment. The O($n^2$) complexity for proximity checks is acceptable for small agent populations ($n < 20$) typical in multi-agent workflow scenarios \cite{qian2023communicative}.

\subsubsection{Theoretical Properties}

The state machine satisfies the following invariants:

\begin{enumerate}
    \item \textbf{Mutual Exclusion}: $\forall a_i: s_i \in \mathcal{S}$ (each agent is in exactly one state)
    \item \textbf{Fatality Irreversibility}: $\forall a_i: s_i = D \Rightarrow \forall t > t_D: s_i = D$ (DECEASED is absorbing)
    \item \textbf{Conservation}: $\sum_{s \in \mathcal{S}} |A_s| = |A|$ (total agent count is preserved)
\end{enumerate}

where $A_s = \{a_i : s_i = s\}$ is the set of agents in state $s$ at time $t$. The model reduces to the classical SIR model when $P_{\text{fat}} = 0$ and $\Delta t_{\text{imm}} = \infty$.
