\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric}
\usepackage{booktabs}

% Use Computer Modern Typewriter (available in BasicTeX) instead of Courier (pcr)
\renewcommand{\ttdefault}{cmtt}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}

\begin{document}

\title{Integrating Cellular Automata with Agent-Based Modeling for Contagion Simulation: A Design Science Research Approach}

\author{\IEEEauthorblockN{Anonymous Author(s)}
\IEEEauthorblockA{\textit{Dept. of Computer Science} \\
\textit{University Name}\\
City, Country \\
email@example.com}
}

\maketitle

\begin{abstract}
The COVID-19 pandemic highlighted a critical accessibility gap in epidemiological modeling tools. Existing agent-based platforms require programming expertise, excluding educators, healthcare professionals, and policy analysts from rapid scenario exploration. We employ Design Science Research methodology to construct a spatial contagion sandbox extending the ChatDev/DevAll multi-agent platform. Our approach integrates cellular automata principles---inspired by Conway's Game of Life---with agent-based modeling, implementing contagion dynamics through local state transitions, proximity-based transmission, and environmental contamination. The system provides zero-code, configuration-driven scenario creation through JSON parameter files. Evaluation demonstrates that simple local transmission rules produce emergent epidemiological patterns: spontaneous quarantine behaviors through collision avoidance, contamination trail dynamics creating persistent transmission reservoirs, and recovery wave propagation. These emergent phenomena validate the cellular automata approach and show that complex macro-level dynamics arise from composable local interactions without centralized coordination. This work contributes: (1) a novel hybrid architecture combining cellular automata with agent-based modeling and LLM-powered systems, (2) design patterns for embedding contagion engines within multi-agent platforms, and (3) a zero-code approach enabling non-programmers to create epidemiological simulations. The artifact democratizes access to contagion modeling for pandemic preparedness training, epidemiology education, and intervention design exploration.
\end{abstract}

\begin{IEEEkeywords}
agent-based modeling, cellular automata, Conway's Game of Life, contagion simulation, Design Science Research, epidemiology, multi-agent systems
\end{IEEEkeywords}

\section{Introduction}
\input{sections/introduction}

\section{Research Methodology}
\input{sections/methodology}

\section{Problem Statement and Research Questions}
\input{sections/problem-statement}

\section{Real-World Applications and Impact}
\input{sections/applications}

\section{Related Work}
\label{sec:related}
\input{sections/related-work-ca}
\input{sections/related-work-epidemiology}

\section{System Design and Implementation}
\label{sec:system-design}
\input{sections/system-architecture}
\input{sections/contagion-formal-spec}
\input{sections/transmission-mechanisms}

\section{Evaluation and Results}
\label{sec:evaluation}
\input{sections/validation-scenarios}
\input{sections/emergent-behavior}

\section{Discussion}
\input{sections/discussion}

\section{Conclusion}
\input{sections/conclusion}

\section*{Acknowledgment}
To be added.

\section{Reproducibility}
\input{sections/reproducibility}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
